---
title: "Gov 50 Final Project"
author: "Layla Chaaraoui"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to the lasting effects of redlining I'd like to explore how in areas that were redlined, what racial populations exist and how these numbers compare to each other. 

## Finding data and writing a proposal 

Data Source: The Lasting Legacy of Redlining, https://github.com/fivethirtyeight/data/tree/master/redlining 

How has redlining impacted Asian populations? In this study, I plan to examine the relationship between redlining and its impact on racial populations, specifically Asian populations. I hypothesize that the `metro_area`s in the dataset that received a low `holc_grade` score will have the lowest amounts of Asian populations then of other demographic groups, such as white and Black. I expect that the Asian population group will be the lowest in comparison to the other demographic groups, even in communities with high scores, but especially in those who received lower scores of C and D, thus receiving this score due to its low Asian population and disproportional community demographics. The study I am evaluating is "The Lasting Legacy of Redlining" by FiveThirtyEight. This study contains 2020 population estimates, as determined by the 2020 U.S. Census, in zones for each redlining grade maps, scored by the Home Owners Loan Corporation (HOLC). My sample is the `metro_area` evaluated, or cities and their state in micro- or metropolitan areas. The explanatory variables of interest I will be studying include `holc_grade` the score the city received from the HOLC, ranging from A (best) to D (hazardous). Additionally, I will be studying the racial demographics of each city/state, which are identified by population numbers (`white_pop`, `black_pop`, `asian_pop`), etc. My outcome variable of interest is the percentage of the populations to identify disproportionate population numbers (`pct_white`, `pct_black`, `pct_asian`), etc. Variable `holc_grade` could also be an outcome variable of interest in comparing grades to Asian population numbers. The observed pattern in the data that would support my hypothesis is if the areas in the study had lower Asian populations in comparison to the other racial demographic groups. Additionally, the populations with disproportionately lower Asian populations will have received lower grades in comparison. A pattern that would disprove my hypothesis is if Asian populations are higher in comparison to other minority groups in the city/states, or if the grades received due to lower Asian populations remain high. 

## One data visualization

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
```

```{r}
metro <- read_csv("data/metro-grades.csv")
```


```{r}
options(scipen = 999)
ggplot(data = metro,
       mapping = aes(x = total_pop,
                     y = pct_asian,
                     color = holc_grade
                     )) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  scale_x_log10() +
  labs(x = "Total Population of Metro Area",
       y = "Percentage of Asian Population Compared to Total Population",
       title = "Comparison Between Asian and Total Population of Metro Areas",
       caption = "Source: FiveThirtyEight",
       color = "HOLC Grade")
```



